# Reproducible Research: Peer Assessment 1
Prepared by Amperio; last version created `r date()`

```{r options_setting, echo=TRUE, results='asis'}
```

## Loading and preprocessing the data
_**Note**: it is assumed that the data has been obtained in its 'original' form, as a compressed file named 'activity.zip' that contains a file named 'activity.csv' with the actual data. The format of this data file is the one presented in the Assignment introduction._

The steps for loading and processing the data are the following:

1. Read the data from the zipped file (using the 'unz' function within the 'read.csv'), with comma as a separator and using the first row for the filed names:

   ```{r}
   data <- read.csv(unz("activity.zip", "activity.csv"), header = TRUE, sep = ",")
   ```

2. Show some records at the start and the end of the data to make sure that the data has been read correctly:

   ```{r}
   head(data)
   tail(data)
   ```

3. As it will be useful later in the analysis, we calculate the average number of steps by 5-minutes intervals accross all the days, and store it as a separate column in the data frame:
   ```{r}
   data.steps_by_interval <- split(data$steps, data$interval)
   data.mean_steps_by_interval <- as.integer(lapply(data.steps_by_interval, mean, na.rm=TRUE))
   data <- cbind(data, data.mean_steps_by_interval)
   colnames(data)[4] <- "avg_steps_in_interval"
   ```

4. It will also be useful to identify each date as a 'weekday' or 'weekend'; we therefore transform the 'date' variable into a separate one indicating the weekday number (1 is Monday, 2 is Tuesday, etc.):
   ```{r}
   data.dates_as_numbers <- strftime(as.Date(data$date), format = "%u")
   data <- cbind(data, data.dates_as_numbers)
   colnames(data)[5] <- "weekday_number"
   ```

5. The final data frame is therefore:
   ```{r}
   head(data)
   tail(data)
   ```

## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
